Mental Rotation as Bayesian Quadrature

### 2013-08-04 man/nips-2013-rebuttal.md

# NIPS 2013 Rebuttal

**Paper ID**: 1352  
**Title**: Mental Rotation as Bayesian Quadrature

## Reviewer 3

Reviewer 3 was primarily concerned with the comparison between models
and human data. While we did not collect human data ourselves, we
compared our model to the human response times reported by Shepard and
Metzler [6] (the key comparison is between Figure 1, depicting the
classic results, and Figure 5, illustrating our modeling results).
The mental rotation results are some of the most famous findings in
cognitive science ([6] alone has at least 2,061 citations according to
Web of Knowledge; 3,640 according to Google Scholar) and have been
replicated many times with many types of stimuli (Thomas 2013). As a
proof-of-concept, we demonstrate that our model is qualitatively
similar to the standard mental rotation result.  No other
computational models have been proposed which can explain how people
choose the correct direction and duration of rotation.

The reviewer was also concerned about how the prior over stimuli
related to the process by which the shapes were created. To clarify,
the shapes were generated in a manner consistent with the prior:
vertex angles were chosen uniformly at random between 0 and $2\pi$,
and vertex radii were chosen uniformly at random between 0 and 1.
Edges were drawn between vertices which were adjacent when sorted by
angle.

## Reviewer 4

Reviewer 4 was concerned with the explanatory power of our model,
suggesting that people do not need to integrate over the full set of
rotations. However, it is not obvious how the strategy "just rotate
until they find a match that is easy to recognize or exhaust the set
of potential rotations" should be implemented. Similarly, stopping
when a local maximum is found is not equivalent to quitting "even if
it's obviously not a match at that rotation". What does it mean for a
match to be easy to recognize, or for something be obviously not a
match? A perfect match cannot truly be identified due to perceptual
uncertainty in people's visual system. Moreover, one cannot just
rotate until a threshold of similarity has been reached, because the
similarity function is not normalized and therefore the value of its
global maximum is not known *a priori*.

If the value of the global maximum of the similarity function is not
known, then it is impossible to know if any particular maximum is the
global maximum or not. To reliably find the global maximum, then,
would require an exhaustive search (a full rotation of 360 degrees),
which is inconsistent with the linear nature of human response times
from [6]. Thus, it is necessary to either be satisfied with a local
maximum (as found by hill-climbing), or by a more sophisticated method.

Even if it was possible to perfectly recognize a match after rotating
the appropriate amount, there is still a separate issue: how do people
know which direction to rotate? Simply choosing one direction at
random and rotating to the true angle produces a pattern of data which
is inconsistent with the behavioral data from [6]. For any true
minimum angle $r$, we would have expected response time proportional
to $E[r] = 0.5r + 0.5(2\pi - r) = \pi$, which is constant rather than
monotonic.

The "model rotations" described around line 372 and shown in Figure 5
do refer to the number of steps the models takes. A more precise label
would be "total distance rotated, in degrees" which is directly
proportional to the number of steps the model takes.

To address the reviewer's question about whether the behavioral
results hold for our stimuli, we emphasize (as above in response to
Reviewer 3) that the Shepard and Metzler results [6] are very robust:
they have been replicated many times and with many other types of
stimuli. In particular, our stimuli are similar to those used by [23].

The reviewer correctly notes that the 95% confidence threshold used by
the BQ model's hypothesis test is likely the source of the 4% error
rate. The reviewer also points out that the MSE on line 382 is
inconsistent with our definition of MSE. This is indeed a mistake: the
MSE cannot be normalized between 0 and 1 because it has no upper
bound. The correct, non-normalized MSEs are 0.03 for the naive model,
0.002 for the parametric model, and 0.001 for the nonparametric model.

## Reviewer 5

We thank Reviewer 5 for their encouraging comments. We do intend to
collect human data in the future and to perform a more careful,
quantitative comparison between the model and participants. We further
hope that the model will be useful in uncovering previously unexamined
nuances in people's behavior.

Regarding people's mental representations, we do not think that people
are necessarily performing full Bayesian quadrature in their
heads. Ours is a computational-level model [3], which means that it is
a formal way of describing how the problem of determining if two
objects are the same could be solved. Although we incorporate the
constraint of sequential rotations, this is not a true
algorithmic-level constraint, as we have not given specific limits on
working memory, processing power, knowledge representation, etc. The
algorithmic-level form of this solution would be an approximation to
our computational-level model and would make the resource limitations
explicit. Currently, however, we have not explored possible
algorithmic approximations. Doing so is a direction for future work.

# References

* Thomas, Nigel J.T., "Supplement: Mental Rotation", *The Stanford
  Encyclopedia of Philosophy*, Edward N. Zalta (ed.). Retrieved on
  08/05/2013 from
  http://plato.stanford.edu/archives/spr2013/entries/mental-imagery/mental-rotation.html

* [6] R. N. Shepard and J. Metzler, "Mental Rotation of
  Three-Dimensional Objects," *Science*, vol. 171, no. 3972,
  pp. 701–703, 1971.

* [23] L. A. Cooper, "Mental Rotation of Random Two-Dimensional
  Shapes," *Cognitive Psychology*, vol. 7, pp. 20-43, 1975.

### 2013-10-09 man/nips-2013-bayesian-optimization.tex
Begin work on bayesian optimization submission

\newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\MSE}[0]{\mathrm{MSE}}
\newcommand{\ME}[0]{\mathrm{ME}}
\newcommand{\naive}[0]{na\"ive}
\newcommand{\Naive}[0]{Na\"ive}

@@

%\nipsfinalcopy % Uncomment for camera-ready version

@@

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
integration over a probability distribution, with the direction of
rotation becoming an optimal experiment design problem (or in machine
learning parlance, an active learning problem). We show that recent
work on methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm} provides a way to
solve this problem, outperforming several simpler heuristics for
determining both the direction and extent of rotation.

The plan of the paper is as follows. First, we give a brief overview
of the work on mental rotation and active learning. Next, we present
the problem domain and a computational-level analysis of how to solve
it. We describe three different models which approximate this
solution, evaluate their accuracy at approximating the computational
solution, and qualitatively compare their behavior to that of the
classic mental rotation results. We end with a brief summary and
conclusion of our approach.

\section{Background}

@@

There is a long and rich history of work on mental rotation beginning
with Shepard and Metzler \cite{Shepard1971}, who presented
participants with pairs of images such as those shown in Figure
\ref{fig:mental-rotation} (left panel). These pairs of
three-dimensional objects could either be: the same object rotated in
the plane (Figure \ref{fig:mental-rotation}A), the same object rotated
in depth (Figure \ref{fig:mental-rotation}B), or two different objects
(Figure \ref{fig:mental-rotation}C; in this case, the ``different''
objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and Shepard and Metzler famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects (Figure \ref{fig:mental-rotation}, right panel). The
conclusion was that participants were visually ``rotating'' the
objects in their minds.

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature
(e.g., \cite{Pylyshyn1981}). In particular, Anderson
\cite{Anderson1978} proved that the mental imagery debate could not be
resolved on the basis of response time data alone. In recent years,
however, significant efforts have investigated the neural
underpinnings of the mental rotation phenomena, arguing that the
results of brain imaging studies support the existence of visual
imagery \cite{Kosslyn:2009tj,Kosslyn1988}.

We assume that mental imagery is indeed a visual process, and turn to
the question of how, and when, it is used. Previous work has examined
how people might use imagery \cite{Hegarty2004, Schwartz1999} to solve
reasoning problems, and when they might use imagery as opposed to a
symbolic rule \cite{Schwartz:1996uy}. Eye-tracking studies of people
performing mental rotations suggest strategies that people might be
using, such as aligning smaller features of the images
\cite{Just1976}. People seem to use imagery methodically, which raises
the question: what is the method?

Others studying people's pattern of behavior when engaged in
self-directed learning have found that they tend to use a strategy of
\textit{active learning} \cite{Gureckis:2012gu,Markant:2012uu}. When
given a choice as to the information they can obtain, people will not
choose the information randomly (as is the case with a \naive{} Monte
Carlo sampler), but according to some utility function such as
information gain \cite{Markant:2012uu,Nelson2007}. We propose that
such a strategy can be combined with the notion of mental imagery as a
tool: an optimal model of mental rotation may, perhaps, rely on an
active learning strategy. In the next sections, we investigate this
idea with several different approaches.

@@

We can formulate the judgment of whether $X_a$ and $X_b$ have the same
origins by deciding about two hypotheses, $h_0$: $\forall M,R\ X_b\neq
MRX_a$ and $h_1$: $\exists M,R\textrm{ s.t. } X_b=MRX_a$.  To compare
the hypotheses, we need to compute the posterior for each:
\begin{equation}
p(h\ \vert\ X_a, X_b)\propto p(X_a, X_b\ \vert\ h)p(h).
\end{equation}
Assuming the hypotheses are equally likely \textit{a priori}, the
prior term $p(h)$ will cancel out when comparing $h_0$ and $h_1$, thus
allowing us to focus on the likelihoods:
\begin{align}
  p(X_a,\ X_b\ \vert \ h_0)&=p(X_a)p(X_b), \label{eq:lh-h0}\\
  p(X_a,\ X_b\ \vert \ h_1)&=\int_R\int_M p(X_a) p(X_b\vert X_a,R,M) p(R) p(M)\ \mathrm{d}M\ \mathrm{d}R. \label{eq:lh-h1}
\end{align}
Under $h_0$, the likelihood is easy to compute because we assume that
$X_a$ and $X_b$ are independent (Equation \ref{eq:lh-h0}). When the
configurations are the same, the likelihood becomes more complicated
(Equation \ref{eq:lh-h1}). For a small number of vertices, we can
compute the integral over $M$ by enumerating every possible mapping
between $X_a$ and $X_b$. After doing so, we obtain:
\begin{equation} 
  p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert X_a,R) p(R)\ \mathrm{d}R.
\end{equation}

Once we have computed both likelihoods, we compute their ratio:
\begin{equation}
  \ell=\frac{p(X_a, X_b\ \vert \ h_1)}{p(X_a, X_b\ \vert \ h_0)}=\frac{\int_R p(X_b\ \vert\ X_a, R)p(R)\ \mathrm{d}R}{p(X_b)}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell<1$, then $h_0$ is the more likely hypothesis. If $\ell>1$,
then $h_1$ is the more likely hypothesis.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/stimuli_shapes.pdf}
    \vspace{0pt}
    \caption{Example stimuli}
    \label{fig:stimuli}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/likelihood_function.pdf}
    \caption{Likelihood function}
    \label{fig:likelihood}
  \end{subfigure}
  \caption{\textbf{Rotated shapes and their similarity.}  \textbf{(a)}
    An example stimulus in which the shapes differ only by a
    rotation. All stimuli consist of three to six vertices centered
    around the origin, and edges which create a closed loop from the
    vertices. \textbf{(b)} The approximate likelihood (similarity) of
    $X_b$ given $X_R$, where $X_R$ is a rotation of $X_a$ by the angle
    $R$. The true angle of rotation between $X_a$ and $X_b$ is at
    $\frac{2\pi}{3}$, corresponding to the global maximum in the
    likelihood function.}
  \label{fig:shapes}
\end{figure}

Given the computational model defined above, we now explore the ways
in which it can be expressed algorithmically. First, we define the
prior probabilities over stimuli according to a generative
procedure. A set of $n$ vertices could be chosen in any of $n!$
different ways, and each vertex is located at a random angle (between
0 and $2\pi$) and radius (between 0 and 1). Thus, the prior over a
shape $X$ is $p(X)=n!\left(\frac{1}{2\pi}\right)^n$, which gives us
the denominator in Equation \ref{eq:lh-ratio}. Computing the numerator
is more difficult, as we we cannot compute $p(X_b\vert X_a, R)$
directly. Instead, we introduce a new variable $X_R$ denoting a mental
image, which approximates $RX_a$. The $X_R$ are generated by repeated
application of a function $\tau$:
\begin{align}
  X_R=RX_a&=\tau(X_{R-r}, r)=\tau(\tau(X_{R-2r}, r), r)=\ldots{}\nonumber \\
  &=\tau^{(\frac{R}{r})}(X_a, r).
  \label{eq:tau}
\end{align} 
where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  p(X_a, X_b\ \vert \ h_1)&=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}
However, the exact form of $p(X_b\vert X_R)$ is still unknown. We
approximate it with a similarity function $S(X_b, X_R)$, and denote
the resulting integral as $Z$:
\begin{equation}
Z=\int_R S(X_b, X_R)p(R)\ \mathrm{d}R\approx \int_R p(X_b\vert X_R)p(R)\ \mathrm{d}R.
\label{eq:Z}
\end{equation}
We define the similarity function $S$, which incorporates the
different mappings as vertices, as follows. Because the vertices are
connected in a way which forms a closed loop, we need only consider
$2n$ mappings of the $n$ vertices (we assume uncertainty for which is
the ``first'' vertex, and then which of its two neighbors is the
``second''). So, the possible orderings are of the form $M=\lbrace{}0,
1, \ldots{}, n\rbrace{}$, $M=\lbrace{}n, 0, \ldots{}, n-1\rbrace{}$,
and so on. Combining this with a Gaussian similarity metric, we
obtain:
\begin{equation}
  S(X_b, X_R)=\frac{1}{2n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\ \vert \ (MX_R)[i], \Sigma)
  \label{eq:similarity}
\end{equation}
where $i$ denotes the $i^{th}$ vertex. An example stimulus and
corresponding $S$ is shown in Figure \ref{fig:shapes}.

@@

\paragraph{Gold standard}

To compare the accuracy of other models' estimates of $Z$, we computed
a ``gold standard'' by evaluating $S(X_b, X_R)$ at 360 values of $R$
spaced evenly between $0$ and $2\pi$ and estimating the integral using
the trapezoidal rule.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/li_regression.pdf}
    \caption{\Naive{} model}
    \label{fig:li}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/vm_regression.pdf}
    \caption{Parametric model}
    \label{fig:vm}
  \end{subfigure}
  \caption{\textbf{Simple methods of estimating $S$.} In both cases,
    hill-climbing search is used until a maxima is found (in this
    case, at approximately $\frac{\pi}{6}$). The sampled points
    $\mathbf{R}$ (red circles) are then used to estimate $S$ (black
    lines are the true $S$, red lines are the estimate). \textbf{(a)}
    Linear interpolation. The overall estimate of $Z$ here will be too
    low, but certain angles will have a disproportionate contribution
    (e.g., between $\frac{\pi}{4}$ and $\frac{3\pi}{4}$. \textbf{(b)}
    Best fit of scaled Von Mises PDF parameters. As in (a), $Z$ will
    be underestimated. However, the fit around the maximum near
    $\frac{\pi}{6}$ is much more accurate.}
  \label{fig:simple-models}
\end{figure}

@@

\paragraph{Parametric (Von Mises)}

Another strategy is to assume a parametric shape for $S$ and fit the
appropriate parameters. If we assume that $S$ is approximately
unimodal when $h_1$ is true\footnote{This is not always a reasonable
  assumption, see e.g., Figure \ref{fig:vm}.}, then a reasonable form
for the similarity function is that of a ``circular Gaussian'' or Von
Mises distribution:
\begin{equation}
  S(X_b, X_R) \approx h\cdot{}p(R\ \vert\ \hat{\theta}, \kappa)=\frac{h}{2\pi I_0(\kappa)}e^{\kappa\cos(R-\hat{\theta})}
\end{equation}
where $\kappa$ is the concentration parameter, $\hat{\theta}$ is the
preferred direction, $h$ is a scale parameter, and $I_0$ is the
modified Bessel function of order zero. We fit these parameters by
minimizing the mean squared error between this PDF and the computed
values of $S$. To choose the sequence of rotations, we use the same
hill-climbing strategy as in the \naive{} model. Figure \ref{fig:vm}
shows an example of the parametric model's estimate of $S$.

\paragraph{Nonparametric (Bayesian Quadrature)}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/bq_regression.pdf}
  \caption{\textbf{Nonparametric model.} Each panel shows one step of
    the Bayesian Quadrature regression. Upper left: the original
    Gaussian Process (GP) regression for $S$. Lower left: the GP
    regression for $\log(S+1)$. Lower right: the GP regression for
    $\Delta_c=\mu_{\log S} - \log \mu_S$. Upper right: the adjusted
    regression for $S$, where the mean is equal to
    $\mu_S(1+\mu_{\Delta_c})$. The model uses this final estimate to
    compute $Z$ and will continue rotating until the variance of $Z$
    is low enough that a hypothesis may be accepted. This method
    allows the model to avoid local maxima such as the one near
    $\frac{\pi}{6}$, which causes trouble for the \naive{} and
    parametric models in Figure \ref{fig:simple-models}.}
  \label{fig:bq}
\end{figure}

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} \cite{Diaconis:1988uo,OHagan:1991tx} to estimate $Z$.
Bayesian Quadrature allows us to compute a posterior distribution over
$Z$ by placing a Gaussian Process (GP) prior on the function $S$ and
evaluating $S$ at a particular set of points. Because our data is
circular, we use a periodic kernel \cite{Rasmussen:2006vz}:
\begin{equation}
k(R, R^\prime)=h^2\exp\left(-\frac{2\sin^2\left(\frac{1}{2}(R-R^\prime)\right)}{w^2}\right).
\end{equation}

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method to place a prior over the log
likelihood\footnote{In practice, as in \cite{Osborne:2012tm}, we use
  the transform of $\log(S+1)$. Additionally, we follow
  \cite{Osborne:2012tm} and use a combination of MLII and
  marginalization to fit the $w$ kernel parameters. The output scales
  are fixed at $h_S=\sqrt{0.1}$ and $h_{\log S}=\log(h_S + 1)$.}, thus
ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points} $c$ as described in \cite{Osborne:2012tm}. The variance is
$\tilde{V}(Z\vert S, \log S, \Delta_c)$ as defined in Equation 12 of
\cite{Osborne:2012tm}.

>>>>

\newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\ME}[0]{\mathrm{ME}}
\newcommand{\naive}[0]{na\"ive}
\newcommand{\Naive}[0]{Na\"ive}

@@

\nipsfinalcopy % Uncomment for camera-ready version

@@

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
integration over a probability distribution, with the direction of
rotation becoming an optimal experiment design problem (or in machine
learning parlance, an active learning problem). We show that recent
work on methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm} provides a way to
solve this problem, outperforming simpler heuristics for determining
the direction and extent of rotation.

The plan of the paper is as follows. First, we present the problem
domain and a computational-level analysis of how to solve it. We
describe two different models which approximate this solution,
evaluate their accuracy at approximating the computational solution,
and qualitatively compare their behavior to that of the classic mental
rotation results. We end with a brief summary and conclusion of our
approach.

@@

% \section{Background}

% There is a long and rich history of work on mental rotation beginning
% with Shepard and Metzler \cite{Shepard1971}, who presented
% participants with pairs of images such as those shown in Figure
% \ref{fig:mental-rotation} (left panel). These pairs of
% three-dimensional objects could either be: the same object rotated in
% the plane (Figure \ref{fig:mental-rotation}A), the same object rotated
% in depth (Figure \ref{fig:mental-rotation}B), or two different objects
% (Figure \ref{fig:mental-rotation}C; in this case, the ``different''
% objects had reflective symmetry, but no rotational
% symmetry). Participants had to determine whether the two images were
% of the same object or not, and Shepard and Metzler famously found that
% response times for plane and depth rotations had a strong linear
% relationship with the minimum angle of rotation between the two
% objects (Figure \ref{fig:mental-rotation}, right panel). The
% conclusion was that participants were visually ``rotating'' the
% objects in their minds.

% For many years, this idea was contested, with some researchers arguing
% the underlying cognitive processes were not visual in nature (e.g.,
% \cite{Pylyshyn1981}). In particular, Anderson \cite{Anderson1978}
% proved that the mental imagery debate could not be resolved on the
% basis of response time data alone. In recent years, however,
% significant efforts have investigated the neural underpinnings of the
% mental rotation phenomena, arguing that the results of brain imaging
% studies support the existence of visual imagery
% \cite{Kosslyn:2009tj,Kosslyn1988}.  We assume that mental imagery is
% indeed a visual process, and turn to the question of how, and when, it
% is used. Previous work has examined how people might use imagery
% \cite{Hegarty2004, Schwartz1999} to solve reasoning problems, and when
% they might use imagery as opposed to a symbolic rule
% \cite{Schwartz:1996uy}. Eye-tracking studies of people performing
% mental rotations suggest strategies that people might be using, such
% as aligning smaller features of the images \cite{Just1976}. People
% seem to use imagery methodically, which raises the question: what is
% the method?

% Others studying people's pattern of behavior when engaged in
% self-directed learning have found that they tend to use a strategy of
% \textit{active learning} \cite{Gureckis:2012gu,Markant:2012uu}. When
% given a choice as to the information they can obtain, people will not
% choose the information randomly (as is the case with a \naive{} Monte
% Carlo sampler), but according to some utility function such as
% information gain \cite{Markant:2012uu,Nelson2007}. We propose that
% such a strategy can be combined with the notion of mental imagery as a
% tool: an optimal model of mental rotation may, perhaps, rely on an
% active learning strategy. In the next sections, we investigate this
% idea with several different approaches.

@@

We can formulate the judgment of whether $X_a$ and $X_b$ have the same
origins by deciding about two hypotheses, $h_0$: $\forall M,R\ X_b\neq
MRX_a$ and $h_1$: $\exists M,R\textrm{ s.t. } X_b=MRX_a$.  To compare
the hypotheses, we need to compute the posterior for each: $p(h\
\vert\ X_a, X_b)\propto p(X_a, X_b\ \vert\ h)p(h)$.

Assuming the hypotheses are equally likely \textit{a priori}, the
prior term $p(h)$ will cancel out when comparing $h_0$ and $h_1$, thus
allowing us to focus on the likelihoods:
\begin{align}
  p(X_a,\ X_b\ \vert \ h_0)&=p(X_a)p(X_b), \label{eq:lh-h0}\\
  p(X_a,\ X_b\ \vert \ h_1)&=\int_R\int_M p(X_a) p(X_b\vert X_a,R,M) p(R) p(M)\ \mathrm{d}M\ \mathrm{d}R. \label{eq:lh-h1}
\end{align}
Under $h_0$, the likelihood is easy to compute because we assume that
$X_a$ and $X_b$ are independent (Equation \ref{eq:lh-h0}). When the
configurations are the same, the likelihood becomes more complicated
(Equation \ref{eq:lh-h1}). For a small number of vertices, we can
compute the integral over $M$ by enumerating every possible mapping
between $X_a$ and $X_b$. After doing so, we obtain:
\begin{equation} 
  p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert X_a,R) p(R)\ \mathrm{d}R.
\end{equation}

Once we have computed both likelihoods, we compute their ratio $\ell$
which is given by $\ell=\left(\int_R p(X_b\ \vert\ X_a, R)p(R)\
  \mathrm{d}R\right) /\ p(X_b)$. If $\ell<1$, then $h_0$ is the more
likely hypothesis. If $\ell>1$, then $h_1$ is the more likely
hypothesis.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/stimuli_shapes.pdf}
    \vspace{0pt}
    \caption{Example stimuli}
    \label{fig:stimuli}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/li_regression.pdf}
    \caption{\Naive{} model}
    \label{fig:li}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/bq_regression_final.pdf}
    \caption{Bayesian Quadrature Model}
    \vspace{0pt}
    \label{fig:bq}
  \end{subfigure}
  \caption{\textbf{(a)} An example stimulus in which the shapes differ
    only by a rotation. All stimuli consist of three to six vertices
    centered around the origin, and edges which create a closed loop
    from the vertices. The true angle of rotation between $X_a$ and
    $X_b$ is at $\frac{2\pi}{3}$. \textbf{(b-c)} Likelihood function
    and \naive{} (b) and BQ (c) model estimates. The sampled points
    $\mathbf{R}$ (red circles) are then used to estimate $S$ (black
    lines are the true $S$, red lines are the estimate).}
  \label{fig:shapes}
\end{figure}

@@

Given the computational model defined above, we now explore the ways
in which it can be expressed algorithmically. First, we define the
prior probabilities over stimuli according to a generative
procedure. A set of $n$ vertices could be chosen in any of $n!$
different ways, and each vertex is located at a random angle (between
0 and $2\pi$) and radius (between 0 and 1). Thus, the prior over a
shape $X$ is $p(X)=n!\left(\frac{1}{2\pi}\right)^n$, which gives us
the denominator of $\ell$. Computing the numerator is more difficult,
as we we cannot compute $p(X_b\vert X_a, R)$ directly. Instead, we
introduce a new variable $X_R$ denoting a mental image, which
approximates $RX_a$. The $X_R$ are generated by repeated application
of a function $\tau$:
\begin{equation}
  X_R=RX_a=\tau(X_{R-r}, r)=\tau(\tau(X_{R-2r}, r), r)=\ldots{}=\tau^{(\frac{R}{r})}(X_a, r).
  \label{eq:tau}
\end{equation} 
where $r$ is a small angle, and $\tau^{(i)}$ indicates $i$ recursive
applications of $\tau$. Using this sequential function, we get:
\begin{align}
  p(X_a, X_b\ \vert \ h_1)&=\int_R \int_{X} p(X_b\vert X) p(X\vert X_a, R)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R \int_X p(X_b\vert X)\delta(\tau^{(\frac{R}{r})}(X_a, r)-X)p(X_a)p(R)\ \mathrm{d}X\ \mathrm{d}R \nonumber \\
  &= \int_R p(X_b\vert X_R)p(X_a)p(R)\ \mathrm{d}R
\end{align}
However, the exact form of $p(X_b\vert X_R)$ is still unknown. We
approximate it with a similarity function $S(X_b, X_R)$, and denote
the resulting integral as $Z$:
\begin{equation}
Z=\int_R S(X_b, X_R)p(R)\ \mathrm{d}R\approx \int_R p(X_b\vert X_R)p(R)\ \mathrm{d}R.
\label{eq:Z}
\end{equation}
We define the similarity function $S$, which incorporates the
different mappings as vertices, as follows. Because the vertices are
connected in a way which forms a closed loop, we need only consider
$2n$ mappings of the $n$ vertices (we assume uncertainty for which is
the ``first'' vertex, and then which of its two neighbors is the
``second''). So, the possible orderings are of the form $M=\lbrace{}0,
1, \ldots{}, n\rbrace{}$, $M=\lbrace{}n, 0, \ldots{}, n-1\rbrace{}$,
and so on. Combining this with a Gaussian similarity metric, we
obtain:
\begin{equation}
  S(X_b, X_R)=\frac{1}{2n}\sum_{M\in\mathbb{M}}\prod_{i=1}^n\mathcal{N}(X_b[i]\ \vert \ (MX_R)[i], \Sigma)
  \label{eq:similarity}
\end{equation}
where $i$ denotes the $i^{th}$ vertex. An example stimulus and
corresponding $S$ is shown in Figure \ref{fig:shapes}.

@@

@@

@@

\paragraph{Bayesian Quadrature}

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} \cite{Diaconis:1988uo,OHagan:1991tx} to estimate $Z$.
Bayesian Quadrature (BQ) allows us to compute a posterior distribution
over $Z$ by placing a Gaussian Process (GP) prior on the function $S$
and evaluating $S$ at a particular set of points. Because our data is
circular, we use a periodic kernel \cite{Rasmussen:2006vz}.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

BQ has its difficulties, however. While in our case $S$ is a
non-negative likelihood function, GP regression enforces no such
constraint. In an effort to avoid this problem, \cite{Osborne:2012tm}
give a method to place a prior over the log likelihood\footnote{In
  practice, as in \cite{Osborne:2012tm}, we use the transform of
  $\log(S+1)$. Additionally, we follow \cite{Osborne:2012tm} and use a
  combination of MLII and marginalization to fit the $w$ kernel
  parameters. The output scales are fixed at $h_S=\sqrt{0.1}$ and
  $h_{\log S}=\log(h_S + 1)$.}, thus ensuring that $S=e^{\log S}$ will
be positive:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points} $c$ as described in \cite{Osborne:2012tm}. The variance is
$\tilde{V}(Z\vert S, \log S, \Delta_c)$ as defined in Equation 12 of
\cite{Osborne:2012tm}.

To start, we pick an initial direction of rotation which results in
the higher value of $S$. At each step, we compute $\mu_Z$ and
$\tilde{V}$ in order to estimate a distribution over the likelihood
ratio $\ell$, i.e.  $p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\
\vert\ \mu_Z, \sigma_z)$.  We choose $h_0$ when $p(\ell < 1)\geq
0.95$, and $h_1$ when $p(\ell > 1)\geq 0.95$. Until one of these
conditions are met (or the shape has been fully rotated), the model
will continues to compute rotations and update its estimate of $Z$.
